% arara: pdflatex: { shell: yes, interaction: nonstopmode }
% arara: pythontex: {verbose: yes, rerun: modified }
% arara: pdflatex: { shell: yes, interaction: nonstopmode }
% arara: clean: { extensions: [ aux, blg, idx, ilg, ind, log, out, pytxcode, rel, toc ] }
% !arara: clean: { files: [ ans.tex, hint.tex] }
\input{header}
\section{Kendall's Notation}
\label{sec:kendalls-notation}

As is apparent from~\cref{sec:constr-discr-time,sec:constr-gg1-queu}, the construction of a queueing process for a single station involves three main elements: the distribution of job inter-arrival times, the distribution of the service times, and the number of servers present to process jobs.



To characterize the type of queueing process it is common to use
\recall{Kendall's abbreviation} $A/B/c/K$, where $A$ is the distribution of the
iid inter-arrival times, $B$ the distribution of the iid service times, $c$ the
number of servers, and $K$ the system size, i.e., the total number of customers that can be simultaneously present, whether in queue or in service.\sidenote{The meaning of $K$ differs among authors. Sometimes it stands for
 the capacity of the queue, not the entire system. In this book $K$ corresponds to the system's size.}
In this notation, it is assumed that the  service discipline is FIFO (First in first out)\sidenote{FCFS (First come first serve) is an other much used accronym for this service discipline.}, i.e., jobs are served in the order in which they arrive.


The most important inter-arrival and service distributions are the exponential distribution, denoted with the shorthand\sidenote{which stands for Markov or Memoryless}  $M$, and the general distribution\sidenote{with the implicit assumption that the first moment is finite},  denoted with  $G$. We write $D$ for a deterministic (constant) random variable.

\newthought{A few important examples} are the following queueing processes: $M/M/1$ in which interarrival times and service times are exponentially distributed;  $M/G/1$ with expoential interarrival times and general service times; and $G/G/c$ in which the interarrival and service times are not specified.

A model that is often used to determine the number of beds needed in (a ward of) a hospital is the $M/M/c/c$ queue.
The motivation is as follows.
Practice tells us that patient inter-arrival times are memoryless, hence exponentially distributed.
Data of patients treatment times shows that these times are often also well-described by an exponential distribution.
Next, there are $c$ beds available, and each bed can serve one patient. When all beds are occupied, the hospital is `full'.


\newthought{When at an} arrival epoch more than one job arrive simultaneously (like a group of people entering a restaurant), we say that a \emph{batch of jobs} arrives.
Likewise, the server can work in batches, for instance, when an oven processes multiple jobs at the same time.
We write $A^X/B^Y/c$ to indicate that we consider batch arrivals and batch services, and then $X$  stands for the distribution of the size of the arriving batch, and $Y$ for the size of the service batch.
When $X\equiv 1$ or $Y \equiv 1$, i.e., single batch arrivals or single batch services, we suppress the $X$ or $Y$ in the queueing formula.

\input{trailer}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
