% arara: pdflatex
% arara: clean: { extensions: [ aux, blg, idx, ilg, ind, log, out, pytxcode, rel, toc ] }
% !arara: clean: { files: [ ans.tex, hint.tex] }

\input{header}

\section{N-policies for the $M/G/1$ queue}
\label{sec:n-policies-mg1}

Interestingly, we can extend the analysis of~\cref{sec:distr-queue-length} and~\cref{sec:n-policies} to compute the average cost of the $M/G/1$ queue under an $N$-policy.
At the end we will find the threshold $N^*$ that minimizes the long-run average cost under the $N$-policy.

Throughout we consider the $M/G/1$ queueing process at moments at which services start.
This is similar to~\cref{sec:distr-queue-length}: When the system is not empty, these are departure epochs.


\newthought{It is easy} to obtain an expression for the clearing time $T(q)$ when a job starts with $q$ jobs in the system.
As in~\cref{sec:distr-queue-length}, write $Y$ for the number of jobs that arrive during a service time.
Then, analogous to~\cref{eq:92}, $T(q)$ satisfies the relation
\begin{equation}\label{eq:33}
  T(q) = \E S + \E{T(q+Y-1)},
\end{equation}
because first the job in service must leave, and then,  when $Y=k$,  it takes $T(q+k-1)$ to hit level~$0$.

Again, we guess that $T(q) = a q + b$.
But $b=0$ since $T(0)=0$.
Substitution of $a q$ into~\cref{eq:33} gives $q a =\E S + aq + a\E Y - a$. And noting that $\E Y = \lambda \E S,$\sidenote{\cref{ex:f-3}} we  obtain
\begin{equation*}
a = \frac{\E S}{1-\lambda \E S} \implies T(q) = \frac{\E S}{1-\lambda \E S}q.
\end{equation*}

\newthought{To find the} cost to clear the queue, we first concentrate on the expected queueing cost $H(q)$ that accrue during the service time of the job.
This consists of two components.
The first is the expected cost of the $q$ jobs in the system while the job is in service.
Clearly, this is $h q \E S$.
The second component is the cost of new jobs that arrive during the service.
While this is slightly harder to determine, we can combine the ideas underlying the derivation of~\cref{eq:99} with \cref{eq:32}.

When a job arrives during a service with remaining time $s$, it has to pay $h s$ directly upon arrival.
Let $U(s)$ denote the expected amount paid by all jobs that arrive during a service time $s$.
If during an interval of length $\delta$, $0<\delta\ll 1$, no new jobs arrive, the expected amount paid is $U(s-\delta)$.\sidenote{Because jobs pay at the moment they arrive.}
If one job arrives, the total amount must be $U(s-\delta) + hs$, and so on.
With probability $1-\lambda \delta$ no job arrives, with probability $\lambda \delta$ one job arrives, and since $\delta\ll 1$, we neglect the probability of multiple arrivals.
Therefore,
\begin{equation*}
  U(s) =  (1-\lambda \delta)\cdot 0 + \lambda \delta \cdot h s + U(s-\delta) + o(\delta).
\end{equation*}
By subtracting $U(s-\delta)$ from both sides, dividing by $\delta$, and taking the limit $\delta \to 0$, this becomes the differential equation $U'(s) = \lambda  h s$.\sidenote{Assuming that $U(\cdot)$ is differentiable.} As $U(0)=0$,  we get that $U(s)= \lambda h s^2/2$.

Since $U(s)$ is the expected total cost given that the service time is~$s$, $\E{U(S)} = \lambda h \E{S^2}/2$ is the expected queueing cost of the arrivals during a service time.

By combining the first and second component of $H(q)$, we obtain
\begin{equation*}
  H(q) = h q \E S + \E{U(S)} = h q \E S + \frac 12 \lambda h \E{S^2}.
\end{equation*}


\newthought{Let $V(q)$ be} the expected queueing costs to clear the system just after a service starts and  $q$ jobs are in the system.
By analogy with~\cref{eq:33}, $V(q)$ must be the solution of
\begin{equation}  \label{eq:98}
  V(q) = H(q) + \E{V(q+Y-1)}.
\end{equation}
Now note that as in~\cref{eq:93}, the cost $H(q)$ has a term linear in~$q$ and a constant term.
As before, we substitute the form $V(q) = aq^2 + bq+c$, assemble terms with the same power in $q$, and solve for the coefficients.
After some work we arrive at\sidenote{~\cref{ex:68}--\cref{ex:nm-2}}
\begin{align*}
  V(q) = \frac{h}{2}\frac{\E S}{1-\rho} q^2 + h  \frac{ 1+ \rho C_s^2}2 \frac{\E S}{(1-\rho)^2} q.
\end{align*}

\newthought{Now that we} have the expected clearing time and cost, we can compute the long-run average cost under a general $N$-policy.
The cycle time $C(N) = N/\lambda + T(N)$, and the queueing cost $W(N)$ until level $N$ is reached while the server is idle is given by~\cref{eq:99}.
Combining all this results in the long-run average cost\sidenote{~\cref{ex:n-mg3}}
\begin{equation}  \label{eq:100}
    \frac{W(N) + K + V(N)}{C(N)}
    = h \frac{1+ C_s^2}2 \frac{\rho^2}{1-\rho} + h \rho + h \frac{N-1}2 + K \frac{\lambda(1-\rho)}N.
\end{equation}
From this, the PK-formula follows directly.\sidenote{~\cref{ex:69}}

\newthought{minimizing over $N$} gives that
\begin{equation*}
  N^* \approx \sqrt{\frac{2\lambda(1-\rho)K}{h}}.
\end{equation*}

The expression for $N^*$, called the \emph{Economic Production Quantity (EPQ)}, is a famous result in inventory theory.
Specifically, it is the optimal production quantity for a production-inventory system in which a machine switches on at cost $K$ and items pay holding cost $h$ per unit time.
By taking $\E S \to 0$, hence $\rho \to 0$, we see that $N^*$ reduces to the \emph{Economic Order Quantity (EPQ)} $\sqrt{2\lambda K/h}$.





\begin{exercise}
Explain intuitively that the system rate-stable for any $N$.
\begin{solution}
  When we switch on the server, the queue `drains' at rate $\mu-\lambda>0$, with $\mu=1/\E S$.
  Consequently, no matter how large $N$, $T(N)<\infty$.
  And, whenever the system is empty, the stochastic process restarts.
  As such cycles start over and over again, and the queue length can never `escape to infinity'.
\end{solution}
\end{exercise}


\begin{exercise}
  Why does the utilization $\rho$ not depend on $N$?
\begin{hint}
 Use the argumentation that leads to~\cref{eq:86}.
\end{hint}
\begin{solution}
  The total number $A(t)$ of job that arrive during $[0,t]$ does not depend on $N$.
  Thus, in~\cref{eq:86}, $\sum_{k=1}^{A(t)}S_k$ does not depend on $N$.
  Now use rate-stability.
\end{solution}
\end{exercise}

\begin{exercise}\label{ex:68}
Simplify $a q^2 +b q = a\E{(q+Y-1)^2} + b\E{q+Y-1}+ H(q)$, and assemble powers in $q$ to obtain:
\begin{align*}
  a &= \frac h 2 \frac{\E S}{1-\E Y} = \frac{h}{2} \frac{\E S}{1- \rho}, \\
  b(1-\E Y) &= a(\E{Y^2} - 2 \E Y + 1) + \frac 12 h \lambda \E{S^2}.
\end{align*}
\begin{hint}
\begin{align*}
  a q^2 &= a q^2, \\
  b q &= 2a q \E Y - 2a q + b q + h q \E S,\\
  0 &= a \E{Y^2} - 2a \E Y + a + b \E Y - b + \frac 12 \lambda h \E{S^2}.
\end{align*}
\end{hint}
\begin{solution}
  In the hint, the first equation is superfluous.
  In the second, $bq$ cancels at both sides, by which we find $a$.
  The third now follows.
\end{solution}
\end{exercise}

\begin{exercise}\label{ex:nm-2}
Derive
\marginpar{In a sense, this is trivial, as it is just algebra, but it is hard to get the details right.}
 the expression for $V(q)$ with the previous exercise.
\begin{hint}
Use~\cref{ex:f-3} to see that $\E{Y^2} = \lambda^2 \E{S^2} + \lambda \E S$.
\end{hint}
\begin{solution}
  For $b$, using the expressions for $\E Y$ and $\E{Y^2}$,
\begin{align*}
b(1-\E Y) &= a(\E{Y^2} - 2 \E Y + 1) + \frac 12 h \lambda \E{S^2} \\
&= \frac{h \E S}{2(1-\E Y)} (\E{Y^2} - 2 \E Y + 1) + \frac 12 h \lambda \E{S^2} \\
&= \frac{h \E S}{2(1-\lambda \E S)} \left(\lambda^2 \E{S^2} + \lambda \E S - 2 \lambda \E S + 1 +  \lambda \E{S^2}\frac{1-\lambda \E S}{\E S}\right) \\
&= \frac{h \E S}{2(1-\lambda \E S)} \left(\lambda^2 \E{S^2} - \lambda \E S + 1 +  \frac{\lambda \E{S^2}}{\E S} - \lambda^2 \E{S^2}\right) \\
&= \frac{h \E S}{2(1-\lambda \E S)} \left(1+ \frac{\lambda \E{S^2}}{\E S }  - \lambda \E S \right) \\
&= \frac{h \E S}{2(1-\lambda \E S)} \left( 1+ \frac{\lambda \E{S^2}}{\E S }  - \lambda \frac{(\E S)^2}{\E S}\right) \\
&= \frac{h \E S}{2(1-\lambda \E S)} \left( 1+ \lambda \frac{\V{S}}{\E S }\right)\\
&= \frac{h \E S}{2(1-\lambda \E S)} \left( 1+ \lambda \frac{\V{S}}{(\E S)^2 } \E S\right)\\
&= \frac{h \E S}{2(1-\lambda \E S)} \left( 1+ \rho C_s^2\right).
\end{align*}
Divide now both sides by $1-\E Y$.
\end{solution}
\end{exercise}

\begin{exercise}
 Check  that $V(q)$ reduces to that of the $M/M/1$ queue.
\begin{solution}
  For $a$, multiply the numerator and denominator by $\mu=1/ \E S$.
  For $b$, multiply by $\mu^2 = 1/(\E S)^2$, use that $C_s^1=1$ because the service times are exponentially distributed, and note that
  \begin{equation*}
    \frac{1+\rho}{1-\rho} = \frac{\mu + \lambda}{\mu-\lambda}.
  \end{equation*}
\end{solution}
\end{exercise}


\begin{exercise}\label{ex:69}
Derive the PK formula from~\cref{eq:100}.
\begin{hint}
Take  $K=0$ and $N=1$, and realize that the LHS is $h\E \L$.
\end{hint}
\begin{solution}
$h\E \L = h \E\Q + h \E\Ls$. $\E \Q = \lambda \E\W$ and $\E \Ls = \lambda \E S$.
\end{solution}
\end{exercise}


\begin{exercise}\label{ex:n-mg3}
Derive~\cref{eq:100}.
\begin{solution}
  Note first that $C(N) = N(1/\lambda + \E S / (1-\rho)) = N/(\lambda(1-\rho))$. Then,
  \begin{align*}
    \frac{V(N) + K + W(N)}{C(N)}
    &= \left(aN^2 + bN + K + h N(N-1)/2 \lambda\right) \frac{\lambda(1-\rho)}N \\
    &= \frac h 2 \rho N  + \frac h 2 \frac \rho{1-\rho} (1+\rho C_s^2) + \frac h 2 (N-1)(1-\rho) + K \frac{\lambda(1-\rho)}N \\
    &= \frac h 2 \frac \rho{1-\rho} (1+\rho C_s^2) + \frac h 2 (N-1 + \rho) + K \frac{\lambda(1-\rho)}N \\
    &= \frac h 2 \frac \rho{1-\rho} (\rho + \rho C_s^2 + 1 - \rho) + \frac h 2 (N-1 + \rho) + K \frac{\lambda(1-\rho)}N \\
    &= \frac h 2 \frac{\rho^2}{1-\rho} (1+ C_s^2) +\frac h 2 \rho + \frac h 2 (N-1 + \rho) + K \frac{\lambda(1-\rho)}N \\
    &= \frac h 2 \frac{\rho^2}{1-\rho} (1+ C_s^2) + h \rho + h \frac{N-1}2 + K \frac{\lambda(1-\rho)}N.
  \end{align*}
%With the above expressions for $a$ and $b$ the result follows immediately.
\end{solution}
\end{exercise}

\input{trailer}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
